---
title: "Etosha Heights Met Summaries"
author: "Mike Loranty"
output: 
  pdf_document:
---  
  
  
  
```{r, echo=FALSE, results='asis', warning=FALSE, message = FALSE}
# note we're running this from the windows machines
# read in the met data file
library(tidyverse)
library(zoo)
dat <- read.csv("G:/My Drive/Documents/research/giraffe/eh_met_data/eh_met_data_2024-10-29.csv", 
                header = T)

# create site var
site <- data.frame(site =c("Vlakwater", "Witgat", "Serengeti", "Swartposdam", "Oupos", "Lion Lodge"),
                   LoggerID = c(22569, 22572, 22575, 22576, 22583, 22584))

dat <- full_join(dat,site)

# DEFINE FUNCTIONS TO CALCULATE WATER DAY/YEAR
# note water year begins on July 1 in southern hemisphere
# determine water year from timestamp
wy <- function(x)
{
  #ifelse(is.POSIXct(x),,"Input is not Date/Time class")
  ifelse(month(x)<7,year(x),year(x)+1)
}

# calculate water day from timestamp
wd <- function(x)
{
  ifelse(leap_year(x),
         ifelse(month(x)<7,yday(x)+184,yday(x)-182),
         ifelse(month(x)<7,yday(x)+184,yday(x)-181))
}
# aggregate variables of interest to daily temporal resolution for plotting
# air temperature
t.air <- dat %>%
  select(year, yday, site, AirTemp, WindSpeed.m.s., WindDir,SolarRad.wm2.) %>%
  group_by(site,year, yday) %>%
  summarise(AirTemp = mean(AirTemp, na.rm = T),
            wspd = mean(WindSpeed.m.s., na.rm = T)) %>%
  mutate(tmstmp = strptime(paste(as.character(year),
                          as.character(yday),sep=""),format = "%Y%j")) 
  t.air <- t.air %>%
  group_by(site) %>%
  mutate(t.wk = rollapplyr(AirTemp,7, mean, na.rm = T, fill = NA),
         ws.wk = rollapplyr(wspd,7, mean, na.rm = T, fill = NA))



prcp <- dat %>%
  select(year, yday, site, precip.mm.) %>%
  group_by(site,year, yday) %>%
  summarise(precip.mm = sum(precip.mm., na.rm = T)) %>%
  mutate(tmstmp = strptime(paste(as.character(year),
                          as.character(yday),sep=""),
                            format = "%Y%j"))

# calculate mean temp, precip, & wind speed for each site over all days with observations at each one
mt <- t.air %>%
  select(tmstmp, site, AirTemp) %>%
  pivot_wider(names_from = site, values_from = AirTemp) %>%
  drop_na() %>%
  summarise((across(2:7, mean))) %>%
  round(1)

mp <- prcp %>%
  group_by(site) %>%
  select(tmstmp, site, precip.mm) %>%
  pivot_wider(names_from = site, values_from = precip.mm) %>%
  drop_na() %>%
  summarise((across(2:7, sum))) %>%
  round(0)

mw <- t.air %>%
  select(tmstmp, site, wspd) %>%
  pivot_wider(names_from = site, values_from = wspd) %>%
  drop_na() %>%
  summarise((across(2:7, mean))) %>%
  round(1)

m <- full_join(mt, mp) %>%
      full_join(mw)

#m$Var <- c("Air Temperature", "Precipitation", "Windspeed")
m <- data.frame(t(m))
names(m) <- c("Air Temperature", "Precipitation", "Windspeed")
knitr::kable(m)
```
  
  
```{r, echo=FALSE, results='asis', warning=FALSE}
ggplot(data = t.air,
       aes(x = tmstmp, y = t.wk, group = site, color = site)) +
geom_line() +
labs(color = "Site", 
       y = expression(Air~Temperature~(degree*C)), 
       x = NULL)
```
Figure 1. Smoothed time-series of daily mean temperatures for weather stations at Etosha Heights. Data are smoothed using a 7-day moving average, primarily for visualization. 

```{r, echo=FALSE, results='asis', warning=FALSE}
ggplot(data = t.air,
       aes(x = tmstmp, y = ws.wk, group = site, color = site)) +
geom_line() +
labs(color = "Site", 
       y = expression(Windspeed~(m~s^-1)), 
       x = NULL)
```
Figure 2. Smoothed time-series of daily mean windspeed for weather stations at Etosha Heights. As with temperature in Figure 1, data are smoothed using a 7-day moving average, primarily for visualization. 

```{r}
ggplot(data = prcp,
       aes(x = tmstmp, y = precip.mm, group = site, color = site)) +
geom_bar(stat = "identity", position = "dodge")
```
  
  
  