---
title: "Etosha Heights Met Summaries"
author: "Mike Loranty"
output: 
  pdf_document:
---  
  
  
  
```{r, echo=FALSE, results='hide', warning=FALSE, message = FALSE}
# note we're running this from the windows machines
# read in the met data file
library(tidyverse)
library(zoo)
ifelse(Sys.info()['sysname'] == "Darwin",
       dat <- read.csv("/Users/mloranty/Library/CloudStorage/GoogleDrive-mloranty@colgate.edu/My Drive/Documents/research/giraffe/eh_met_data/eh_met_data_2025-06-30.csv"),
       dat <- read.csv("G:/My Drive/Documents/research/giraffe/eh_met_data/eh_met_data_2025-06-30.csv"))


# create site var
site <- data.frame(site =c("Vlakwater", "Witgat", "Serengeti", "Swartposdam", "Oupos", "Lion Lodge"),
                   loggerID = c("z6-22569", "z6-22572", "z6-22575", "z6-22576", "z6-22583", "z6-22584"))

dat <- full_join(dat,site)

# DEFINE FUNCTIONS TO CALCULATE WATER DAY/YEAR
# note water year begins on July 1 in southern hemisphere
# determine water year from timestamp
wy <- function(x)
{
  #ifelse(is.POSIXct(x),,"Input is not Date/Time class")
  ifelse(month(x)<7,year(x),year(x)+1)
}

# calculate water day from timestamp
wd <- function(x)
{
  ifelse(leap_year(x),
         ifelse(month(x)<7,yday(x)+184,yday(x)-182),
         ifelse(month(x)<7,yday(x)+184,yday(x)-181))
}
# aggregate variables of interest to daily temporal resolution for plotting
# air temperature
t.air <- dat %>%
  select(year, yday, site, t.air, wind.spd, wind.dir,solar.rad) %>%
  distinct() %>%
  group_by(site,year, yday) %>%
  summarise(t.air = mean(t.air, na.rm = T),
            wind.spd = mean(wind.spd, na.rm = T)) %>%
  mutate(tmstmp = strptime(paste(as.character(year),
                          as.character(yday),sep=""),format = "%Y%j")) 
  t.air <- t.air %>%
  group_by(site) %>%
  mutate(t.wk = rollapplyr(t.air,7, mean, na.rm = T, fill = NA),
         ws.wk = rollapplyr(wind.spd,7, mean, na.rm = T, fill = NA))



prcp <- dat %>%
  select(year, yday, site, precip) %>%
  distinct() %>%
  group_by(site,year, yday) %>%
  summarise(precip = sum(precip, na.rm = T)) %>%
  mutate(tmstmp = strptime(paste(as.character(year),
                          as.character(yday),sep=""),
                            format = "%Y%j"))

# calculate mean temp, precip, & wind speed for each site over all days with observations at each one
mt <- t.air %>%
  select(tmstmp, site, t.air) %>%
  pivot_wider(names_from = site, values_from = t.air) %>%
  drop_na() %>%
  summarise((across(2:7, mean))) %>%
  round(1)

mp <- prcp %>%
  group_by(site) %>%
  select(tmstmp, site, precip) %>%
  pivot_wider(names_from = site, values_from = precip) %>%
  drop_na() %>%
  summarise((across(2:7, sum))) %>%
  round(0)

mw <- t.air %>%
  select(tmstmp, site, wind.spd) %>%
  pivot_wider(names_from = site, values_from = wind.spd) %>%
  drop_na() %>%
  summarise((across(2:7, mean))) %>%
  round(1)

m <- full_join(mt, mp) %>%
      full_join(mw)

#m$Var <- c("Air Temperature", "Precipitation", "Windspeed")

```
  
  
```{r, echo=FALSE, results='asis', warning=FALSE}
m <- data.frame(t(m))
names(m) <- c("Air Temperature (C)", "Precipitation (mm)", "Windspeed (m/s)")
knitr::kable(m, caption = "Mean meterological conditions across all days with data at every site. Days with missing data at one or more sites were omitted. n = 353")

ggplot(data = t.air,
       aes(x = tmstmp, y = t.wk, group = site, color = site)) +
geom_line() +
labs(color = "Site", 
       y = expression(Air~Temperature~(degree*C)), 
       x = NULL)
```
Figure 1. Smoothed time-series of daily mean temperatures for weather stations at Etosha Heights. Data are smoothed using a 7-day moving average, primarily for visualization. 

```{r, echo=FALSE, results='asis', warning=FALSE}
ggplot(data = t.air,
       aes(x = tmstmp, y = ws.wk, group = site, color = site)) +
geom_line() +
labs(color = "Site", 
       y = expression(Windspeed~(m~s^-1)), 
       x = NULL)
```
Figure 2. Smoothed time-series of daily mean windspeed for weather stations at Etosha Heights. As with temperature in Figure 1, data are smoothed using a 7-day moving average, primarily for visualization. 



```{r, echo=FALSE, results='asis', warning=FALSE}

# add water year and water day to precip data
prcp$wy <- wy(prcp$tmstmp)
prcp$wd <- wd(prcp$tmstmp)

prcp <- prcp %>% 
  group_by(site, wy) %>% 
  arrange(tmstmp) %>% 
  mutate(precip.cum = cumsum(precip))

prcp %>%
  filter(wy==2024) %>%
ggplot(
       aes(x = tmstmp, y = precip.cum, group = site, color = site)) +
geom_line() +
  labs(color = "Site", 
       y = "Cumulative Precipitation (mm)" , 
       x = NULL)
```
  
Figure 3. Cumulative precipitation by site for the 2024 water year (beginning 1 July 2023). Note there was a sensor outage at Oupos beginning in March 2024 after which no data were recorded.  
  